---
title: "input"
author: "Pierre Nouvellet"
date: "2019-07"
---

```{r options, include = FALSE, message = FALSE, warning = FALSE, error = FALSE}
library(knitr)
library(Hmisc)
library(EpiEstim)
library(zoo)
library(latex2exp)

opts_chunk$set(collapse = TRUE)

path_s <- '../../../../Dropbox (SPH Imperial College)/Mobility.2.0.Save/Saved_file/'

opts_chunk$set(fig.path=paste0(path_s,'inputProcessing/figs/'), fig.keep='high', 
               dev=c('png'), fig.width=9, fig.height=6, cache=FALSE,
               tidy=FALSE, warning=FALSE, fig.show="asis"
 	       )

sapply(paste0('../Rfunction_library/pre_MCMC/',
              (list.files('../Rfunction_library/pre_MCMC/'))),FUN = source)

```

# input data

## load Epi and mobility data

```{r}
# read SI
d0 <- readRDS(file=paste0(path_s,'/inputs/data_2020-09-13.rds'))
mu_si <- d0$si_mean
std_si <- d0$si_std
rm(d0)

# mean sd of ditrivution infection to death
mu_id = 18.8
si_std = .45*mu_id
# si <- 2

```


# read incidence of death and eliminate those with no consecutive 4 weeks with >70 deaths

```{r}
d <- readRDS(file=paste0(path_s,'/inputs/latest_deaths_wide_no_filter.rds'))
col_n <- names(d)
names(d)[which(col_n %in% 'Dominican_Republic')] <- 'Dominican_R.'
names(d)[which(col_n %in% 'United_States_of_America')] <- 'USA'




d$dates <- as.character(d$dates)
d <- d[-c(1:which(d$dates %in% '2020-01-05')),]

daily_to_weekly <- function(x){
  res <- data.frame(dates = seq(x$dates[seq(1,nrow(x),by=7)]),
                    matrix(NA,nrow = nrow(x)/7,ncol = ncol(d)-1))
  for (i in 2:ncol(x)){
    temp <- x[,i]
    res[,i] <- rowSums(matrix(temp,nrow = nrow(res),ncol = 7,byrow = TRUE))
  }
  names(res) <- names(x)
  return(res)
}
d_weekly <- daily_to_weekly(x = d)


keep <- rep(NA,ncol(d_weekly)-1)
for(i in 2:ncol(d_weekly)){
  temp <- paste0(as.numeric(d_weekly[,i] >= 70),collapse = '')
  x <- grep(pattern = '1111',x = temp)
  if(length(x) == 0){
    keep[i-1] <- 0
  }else if(x ==1 ){
    keep[i-1] <- 1
  }else{
    break
  }
}
# sum(eliminate == 0)
d <- d[,c(1,which(keep ==1) + 1)]
d_weekly <- d_weekly[,c(1,which(keep ==1) + 1)]

country_D <- names(d)[-1]
country_D

```

eliminate country with high within week variation

```{r}
rollD <- apply(d[,2:ncol(d)],2,zoo::rollmean,k=7)
obs <- d[4:(nrow(d)-3),2:ncol(d)]
temp <- as.matrix(abs(obs - rollD)/(rollD+1))
f <- which(obs==0)
temp[f] <- NA
elim <- (which(apply(temp,2,mean,na.rm = TRUE)>1))

d <- d[,-(elim+1)]
d_weekly <- d_weekly[,-(elim+1)]
country_D <- names(d)[-1]
country_D


```


```{r}

rm(i,keep,temp,x,col_n,rollD,obs,elim,f)


```


# load mobility and corss reference countries

```{r, echo = FALSE}

dM <- readRDS(file=paste0(path_s,'/inputs/mobility_raw.rds'))
# apple corrections
dM$apple$region[which(dM$apple$region %in% 'Saudi Arabia')] <- 'Saudi_Arabia'
dM$apple$region[which(dM$apple$region %in% 'South Africa')]<- 'South_Africa'
dM$apple$region[which(dM$apple$region %in% 'United Kingdom')] <- 'United_Kingdom'
dM$apple$region[which(dM$apple$region %in% 'United States')] <- 'USA'
#google corrections
dM$google$region[which(dM$google$region %in% 'Dominican Republic')] <- 'Dominican_R.'
dM$google$region[which(dM$google$region %in% 'Saudi Arabia')] <- 'Saudi_Arabia'
dM$google$region[which(dM$google$region %in% 'South Africa')]<- 'South_Africa'
dM$google$region[which(dM$google$region %in% 'United Kingdom')] <- 'United_Kingdom'
dM$google$region[which(dM$google$region %in% 'United States')] <- 'USA'


# check apple
country_a <- sort(unique((dM$apple$region)))
country_D[-which(country_D %in% country_a)]
f <- which(dM$apple$region %in% country_D)
dM$apple <- dM$apple[f,]

#check google
country_g <- sort(unique(as.character(dM$google$region)))
country_D[-which(country_D %in% country_g)]
f <- which(dM$google$region %in% country_D)
dM$google <- dM$google[f,]

to_eliminate <- intersect(country_D[-which(country_D %in% country_a)],country_D[-which(country_D %in% country_g)])
f <- which(names(d) %in% to_eliminate)
d <- d[,-f]

```


```{r}



# sum(names(dM$apple) != names(dM$apple))

dates_c <- as.character(substr(names(dM$apple[,-c(1:3)]),start = 2, stop = 11))
dates_c <- sub(pattern = '.', replacement = '-', x = dates_c, fixed = TRUE)
dates_c <- sub(pattern = '.', replacement = '-', x = dates_c, fixed = TRUE)
# which(x$dates %in% dates_c)

names(dM$apple) <- c(names(dM$apple)[1:3],dates_c)
names(dM$google) <- c(names(dM$google)[1:3],dates_c)

```


```{r}


M <- list()
#Apple
M$apple <- list()
stream_apple <- sort(unique(dM$apple$transportation_type))
for(i in 1:length(stream_apple)){
  x <- d
  x[,2:ncol(x)] <- NA
  y <- ( dM$apple[which(dM$apple$transportation_type == stream_apple[i]),-c(1,3)] )
  
  temp <- data.frame( dates = names(y)[-1],
                      t(as.matrix(y[,2:ncol(y)])))
  names(temp) <- c('dates',y$region)
  rownames(temp) <- 1:nrow(temp)
  temp$dates <- as.character(temp$dates)
  f <- which(x$dates %in% temp$dates)
  f_c <- which(names(x) %in% names(temp))
  
  x[f,f_c] <- temp
  
  M$apple[[i]] <- x
}
names(M$apple) <- stream_apple


#Google
M$google <- list()
stream_google <- sort(unique(dM$google$transportation_type))
for(i in 1:length(stream_google)){
  x <- d
  x[,2:ncol(x)] <- NA
  y <- ( dM$google[which(dM$google$transportation_type == stream_google[i]),-c(1,3)] )
  
  temp <- data.frame( dates = names(y)[-1],
                      t(as.matrix(y[,2:ncol(y)])))
  names(temp) <- c('dates',y$region)
  rownames(temp) <- 1:nrow(temp)
  temp$dates <- as.character(temp$dates)
  f <- which(x$dates %in% temp$dates)
  f_c <- which(names(x) %in% names(temp))
  
  x[f,f_c] <- temp
  
  M$google[[i]] <- x
}
names(M$google) <- stream_google

```

# get averaged - interpolated compound mobility

```{r}
M_process <- M
f_d <- which(M[[1]][[1]]$dates %in% c('2020-02-17','2020-09-06'))

for(i in 1:2){ # apple/google
  if(i==1){
    stream <- stream_apple
  }else if(i==2){
    stream <- stream_google
  }
  for(j in 1:length(stream)){#stream
    for(k in 2:ncol(M[[i]][[j]])){ #country
      
      x <- M[[i]][[j]][,k]
      if(sum(!is.na(x))<2){
        M_process[[i]][[j]][,k+1] <- x
      }else{
        x2 <- na.approx(x,na.rm = FALSE)
        x_roll <- zoo::rollmean(x2,k=7,align = c("center"))
        # > mean(c(102,103,101, 99,104,106,102))
        # [1] 102.4286
        # > mean(c(103,101, 99,104,106,102,104))
        # [1] 102.7143
        x_roll <- c(rep(NA,3),x_roll,rep(NA,3))
        x_roll[f_d[1]:(f_d[1]+2)] <- x_roll[(f_d[1]+3)]
        x_roll[(f_d[2]-2):(f_d[2])] <-x_roll[(f_d[2]-3)]
        
        M_process[[i]][[j]][,k] <- x_roll
      }
    }
  }
}

# plot(x,pch=16)
# lines(x2,col='blue3',lwd=2)
# lines(x_roll,col='red3',lwd=2)

combine_mob <- function(x){
  y <- x[[1]]
  for(i in 2:ncol(y)){
    temp <- matrix(NA,nrow(y),length(x))
    for (j in 1:length(x)){
      temp[,j] <- x[[j]][,i]
    }
    temp <- apply(temp,1,mean,na.rm=TRUE)
    temp[!is.finite(temp)] <- NA
    y[,i] <-temp
  }
  return(y)
}

M_process[['Combined']] <- list()
M_process[['Combined']][['Apple_c']] <- combine_mob(x = list(M_process$apple$driving,
                                                    M_process$apple$transit,
                                                    M_process$apple$walking)) 



M_process[['Combined']][['Google_c']] <- combine_mob(x = list(M_process$google$grocery_and_pharmacy_percent_change_from_baseline,
                                                    M_process$google$retail_and_recreation_percent_change_from_baseline,
                                                    M_process$google$transit_stations_percent_change_from_baseline,
                                                    M_process$google$workplaces_percent_change_from_baseline)) 

M_process[['Combined']][['AG']] <- combine_mob(x = list(M_process$Combined$Apple_c,
                                                    M_process$Combined$Google_c)) 


         

```

# some plots

```{r}
x_t_lab <- as.Date(c('01-01-2020','01-03-2020',
                     '01-05-2020','01-07-2020',
                     '01-09-2020'),format = '%d-%m-%Y')
layout(matrix(1:6,3,2,byrow = FALSE))
col1 <- c('red3','blue3','chartreuse4','cyan4','darkgoldenrod3','darkmagenta')
d$dates <- as.Date(d$dates,format='%Y-%m-%d')
add_c <- c(FALSE,TRUE)
y_la <- c('','Deaths','','Mobility')


# par(mar = c(bottom, left, top, right)) defauklt: c(5.1, 4.1, 4.1, 2.1)
for(i in 2:ncol(d)){
  n_colf <- ((i %% 2) == 0) +1
  par(mar = c(2, 4, 2, 1))
  plot(d$dates,d[,i],main = names(d)[i],pch=16,cex=.8,
       xlab = '',ylab = y_la[n_colf],bty = 'n',xaxt="n")
  axis(side = 1,at = x_t_lab, labels = rep('',length(x_t_lab)))
  #apple
  c <- 1
  for(k in 1:length(stream_apple)){
    y <- M$apple[[k]][,i]
    y2 <- M_process$apple[[k]][,i]
    
    if((sum(!is.na(y))>0) & (c==1)){
      plot(d$dates,y,pch = 3, col=col1[k],cex=.2,
           main = 'Apple',ylim = c(0,200),
           xlab = '',ylab = y_la[n_colf+2],bty = 'n',xaxt="n")
      axis(side = 1,at = x_t_lab, labels = rep('',length(x_t_lab)))
      lines(d$dates,y2, col=col1[k])
      
      abline(h = 100,lty = 3, col='grey')
      c <- 2
    }else if((sum(!is.na(y))>0) & (c==2)){
      lines(d$dates,y, pch = 3, col=col1[k], cex=.2, type='p')
      lines(d$dates,y2, col=col1[k])
      abline(h = 100,lty = 3, col='grey')
    }
  }
  if(c==1){
    plot(d$dates,rep(1e3,nrow(d)),
         ylim = c(0,200),main = 'Apple',
         xlab = '',ylab = y_la[n_colf+2],bty = 'n',xaxt="n")
      axis(side = 1,at = x_t_lab, labels = rep('',length(x_t_lab)))
  }else{
    # combined ones 
      lines(d$dates,M_process$Combined$Apple_c[,i], col='dodgerblue1',lty = 5,lwd=2)
      lines(d$dates,M_process$Combined$AG[,i], col='grey0',lty = 5,lwd=2)
  }
  
  if(n_colf == 2){
    legend('topleft',legend = c('Dr','Tr','Wk'),
           col = col1[1:3],bty = 'n',lwd=1, seg.len = 1)
  }
  #google
  c <- 1
  for(k in 1:length(stream_google)){
    y <- M$google[[k]][,i]
    y2 <- M_process$google[[k]][,i]
    
    if((sum(!is.na(y))>0) & (c==1)){
      plot(d$dates,y,pch = 3, col=col1[k],cex=.2,
           main = 'Google',ylim = c(0,200),
           xlab = '',ylab = y_la[n_colf+2],bty = 'n')
      lines(d$dates,y2, col=col1[k])
      abline(h = 100,lty = 3, col='grey')
      c <- 2
    }else if((sum(!is.na(y))>0 ) & (c==2)){
      lines(d$dates,y,pch = 3, col=col1[k],cex=.2, type='p')
      lines(d$dates,y2, col=col1[k])
      abline(h = 100,lty = 3, col='grey')
    }
  }
  if(c==1){
    plot(d$dates,rep(1e2,nrow(d)),ylim = c(0,1),
         main = 'Google',
         xlab = '',ylab = y_la[n_colf+2],bty = 'n',xaxt="n")
    axis(side = 1,at = x_t_lab, labels = rep('',length(x_t_lab)))
  }else{
    # combined ones 
      lines(d$dates,M_process$Combined$Google_c[,i], col='dodgerblue1',lty = 5,lwd=2)
      lines(d$dates,M_process$Combined$AG[,i], col= 'grey0',lty = 5,lwd=2)
  }
  
  #legends
  if(n_colf == 2){
    legend('topleft',legend = c('Gr','Pk','Rs','Rt','Tr','Wo'),
           col = col1,bty='n',lwd=1, seg.len = 1)
  }else if(n_colf == 1){
    legend('topleft',legend = c('A or G','C'),
           col = c('dodgerblue1','grey0') ,bty='n',lwd=1,lty=5, seg.len = 2)
  }
}


```









# useful data

```{r}
# correct epi
D <- d
mD <- as.matrix(D[,-1])

cumD <- data.frame(D$dates,apply(D[,-1],2,cumsum))


N_geo <- ncol(D)-1
N_days <- nrow(D)
country <- as.character(d$Country)

```




## process Epi data

load Epi data and serial interval
```{r, echo = FALSE}
## # serial interval estimate used: mean = 3.96, sd =  4.75
## # Teunis et al.

SI <- list()

for (si in c(1,2)){
  
  SI[[si]] <- list()
  SI[[si]]$si_mean <- mu_si[si]
  SI[[si]]$si_std <- std_si[si]
  
  SI[[si]]$SI <- gamma_dist_EpiEstim(si_mu =  mu_si[si], 
                            si_std  = std_si[si], 
                            SItrunc = 30)
  
  plot(seq(0, SI[[si]]$SI$SItrunc), SI[[si]]$SI$dist, type = "h",
       lwd = 10, lend = 1, xlab = "time (days)", ylab = "frequency")
  title(main = "Discrete distribution of the serial interval of COVID-19")
  
}

```

## delay infection to death

```{r, echo = FALSE}

delta_id <- gamma_dist_EpiEstim(si_mu = mu_id , 
                                si_std  = .45*mu_id, 
                                SItrunc = 50)

plot(seq(0, delta_id$SItrunc), delta_id$dist, type = "h",
     lwd = 10, lend = 1, xlab = "time (days)", ylab = "frequency")
title(main = "Discrete distribution of the delay infection to death")

sum(delta_id$dist[1:7])


```

# Useful formatting

see paper

```{r}

# delay death infection matrix
H <- matrix(0,nrow = N_days, ncol = N_days)
for (i in 1:N_days){
  f <- max(c(1,i-delta_id$SItrunc))
  H[i,f:i] <- rev(delta_id$dist)[((delta_id$SItrunc+1)-(i-f)):(delta_id$SItrunc+1)]
  if (i>1) H[i,f:i] <-  H[i,f:i]/sum( H[i,f:i])
}

# matrix of mobility and delayed version
M_process_mat <- list()
# counter <- 1
for(i in 1:length(M_process)){
  for(j in 1:length(M_process[[i]])){
    M_process_mat[[as.character(names(M_process[[i]])[j])]] <- as.matrix(M_process[[i]][[j]][,-1])
  }
}
# reformat mobility raw
M_raw <- list()
# counter <- 1
for(i in 1:length(M)){
  for(j in 1:length(M[[i]])){
    M_raw[[as.character(names(M[[i]])[j])]] <- as.matrix(M[[i]][[j]][,-1])
  }
}
# as.matrix(M_process$$mob_combined_smooth[,-1])
# M_D <- H %*% M


# delay SI death to death matrix
W <- list()
Ot <- list()
for (si in c(1,2)){
  
  W[[si]] <- list()
  W[[si]]$W <- matrix(0,nrow = N_days, ncol = N_days)
  for (i in 1:N_days){
    f <- max(c(1,i-SI[[si]]$SI$SItrunc))
    W[[si]]$W[i,f:i] <- rev(SI[[si]]$SI$dist)[((SI[[si]]$SI$SItrunc+1)-(i-f)):(SI[[si]]$SI$SItrunc+1)]
    # if (i>1) H[i,f:i] <-  H[i,f:i]/sum( H[i,f:i])
  }
  
  # overall infectivity matrix
  Ot[[si]] <- list()
  Ot[[si]]$Ot <- W[[si]]$W %*% mD
  # correct infectivity when >0 case observed but no infectivity
  for(i in 1:N_geo){
     Ot[[si]]$Ot[ which( (D[,i+1]>0) & (Ot[[si]]$Ot[,i] ==0 ) ),i] <-NA
  }
}




```

# save output

```{r}

inputs <- list(D = D,
               M = M,
               M_process = M_process,
               M_process_mat = M_process_mat,
               Ot = Ot,
               W = W,
               H = H,
               SI = SI,
               delta_id = delta_id)

saveRDS(object = inputs, 
        file = paste0(path_s,'inputProcessing/Rdata/inputs.rds'))

```