---
title: "input"
author: "Pierre Nouvellet"
date: "2019-07"
---

```{r options, include = FALSE, message = FALSE, warning = FALSE, error = FALSE}
library(knitr)
library(Hmisc)
library(EpiEstim)
library(zoo)
library(latex2exp)

opts_chunk$set(collapse = TRUE)

path_s <- '../../../../Dropbox (SPH Imperial College)/Mobility.2.0.Save/Saved_file/'

opts_chunk$set(fig.path=paste0(path_s,'inputProcessing/figs/'), fig.keep='high', 
               dev=c('png'), fig.width=9, fig.height=6, cache=FALSE,
               tidy=FALSE, warning=FALSE, fig.show="asis"
 	       )

sapply(paste0('../Rfunction_library/pre_MCMC/',
              (list.files('../Rfunction_library/pre_MCMC/'))),FUN = source)

```

# input data

## load Epi and mobility data

```{r}
# read SI
d0 <- readRDS(file=paste0(path_s,'/inputs/data_2020-09-13.rds'))
mu_si <- d0$si_mean
std_si <- d0$si_std
rm(d0)

# mean sd of ditrivution infection to death
mu_id = 18.8
si_std = .45*mu_id
# si <- 2

```


# read incidence of death and eliminate those with no consecutive 4 weeks with >70 deaths

```{r}
d <- readRDS(file=paste0(path_s,'/inputs/latest_deaths_wide_no_filter.rds'))
col_n <- names(d)
names(d)[which(col_n %in% 'Dominican_Republic')] <- 'Dominican_R.'
names(d)[which(col_n %in% 'United_States_of_America')] <- 'USA'




d$dates <- as.character(d$dates)
d <- d[-c(1:which(d$dates %in% '2020-01-05')),]

daily_to_weekly <- function(x){
  res <- data.frame(dates = seq(x$dates[seq(1,nrow(x),by=7)]),
                    matrix(NA,nrow = nrow(x)/7,ncol = ncol(d)-1))
  for (i in 2:ncol(x)){
    temp <- x[,i]
    res[,i] <- rowSums(matrix(temp,nrow = nrow(res),ncol = 7,byrow = TRUE))
  }
  names(res) <- names(x)
  return(res)
}
d_weekly <- daily_to_weekly(x = d)


keep <- rep(NA,ncol(d_weekly)-1)
for(i in 2:ncol(d_weekly)){
  temp <- paste0(as.numeric(d_weekly[,i] >= 70),collapse = '')
  x <- grep(pattern = '1111',x = temp)
  if(length(x) == 0){
    keep[i-1] <- 0
  }else if(x ==1 ){
    keep[i-1] <- 1
  }else{
    break
  }
}
# sum(eliminate == 0)
d <- d[,c(1,which(keep ==1) + 1)]
d_weekly <- d_weekly[,c(1,which(keep ==1) + 1)]

country_D <- names(d)[-1]
country_D

```

eliminate country with high within week variation

```{r}
rollD <- apply(d[,2:ncol(d)],2,zoo::rollmean,k=7)
obs <- d[4:(nrow(d)-3),2:ncol(d)]
temp <- as.matrix(abs(obs - rollD)/(rollD+1))
f <- which(obs==0)
temp[f] <- NA
elim <- (which(apply(temp,2,mean,na.rm = TRUE)>1))

d <- d[,-(elim+1)]
d_weekly <- d_weekly[,-(elim+1)]
country_D <- names(d)[-1]
country_D


```


```{r}

rm(i,keep,temp,x,col_n,rollD,obs,elim,f)


```


# load mobility and corss reference countries

```{r, echo = FALSE}

dM <- readRDS(file=paste0(path_s,'/inputs/mobility_raw.rds'))
# apple corrections
dM$apple$region[which(dM$apple$region %in% 'Saudi Arabia')] <- 'Saudi_Arabia'
dM$apple$region[which(dM$apple$region %in% 'South Africa')]<- 'South_Africa'
dM$apple$region[which(dM$apple$region %in% 'United Kingdom')] <- 'United_Kingdom'
dM$apple$region[which(dM$apple$region %in% 'United States')] <- 'USA'
#google corrections
dM$google$region[which(dM$google$region %in% 'Dominican Republic')] <- 'Dominican_R.'
dM$google$region[which(dM$google$region %in% 'Saudi Arabia')] <- 'Saudi_Arabia'
dM$google$region[which(dM$google$region %in% 'South Africa')]<- 'South_Africa'
dM$google$region[which(dM$google$region %in% 'United Kingdom')] <- 'United_Kingdom'
dM$google$region[which(dM$google$region %in% 'United States')] <- 'USA'


# check apple
country_a <- sort(unique((dM$apple$region)))
country_D[-which(country_D %in% country_a)]
f <- which(dM$apple$region %in% country_D)
dM$apple <- dM$apple[f,]

#check google
country_g <- sort(unique(as.character(dM$google$region)))
country_D[-which(country_D %in% country_g)]
f <- which(dM$google$region %in% country_D)
dM$google <- dM$google[f,]

to_eliminate <- intersect(country_D[-which(country_D %in% country_a)],country_D[-which(country_D %in% country_g)])
f <- which(names(d) %in% to_eliminate)
d <- d[,-f]

```


```{r}



# sum(names(dM$apple) != names(dM$apple))

dates_c <- as.character(substr(names(dM$apple[,-c(1:3)]),start = 2, stop = 11))
dates_c <- sub(pattern = '.', replacement = '-', x = dates_c, fixed = TRUE)
dates_c <- sub(pattern = '.', replacement = '-', x = dates_c, fixed = TRUE)
# which(x$dates %in% dates_c)

names(dM$apple) <- c(names(dM$apple)[1:3],dates_c)
names(dM$google) <- c(names(dM$google)[1:3],dates_c)

```


```{r}


M <- list()
#Apple
M$apple <- list()
stream_apple <- sort(unique(dM$apple$transportation_type))
for(i in 1:length(stream_apple)){
  x <- d
  x[,2:ncol(x)] <- NA
  y <- ( dM$apple[which(dM$apple$transportation_type == stream_apple[i]),-c(1,3)] )
  
  temp <- data.frame( dates = names(y)[-1],
                      t(as.matrix(y[,2:ncol(y)])))
  names(temp) <- c('dates',y$region)
  rownames(temp) <- 1:nrow(temp)
  temp$dates <- as.character(temp$dates)
  f <- which(x$dates %in% temp$dates)
  f_c <- which(names(x) %in% names(temp))
  
  x[f,f_c] <- temp
  
  M$apple[[i]] <- x
}
names(M$apple) <- stream_apple


#Google
M$google <- list()
stream_google <- sort(unique(dM$google$transportation_type))
for(i in 1:length(stream_google)){
  x <- d
  x[,2:ncol(x)] <- NA
  y <- ( dM$google[which(dM$google$transportation_type == stream_google[i]),-c(1,3)] )
  
  temp <- data.frame( dates = names(y)[-1],
                      t(as.matrix(y[,2:ncol(y)])))
  names(temp) <- c('dates',y$region)
  rownames(temp) <- 1:nrow(temp)
  temp$dates <- as.character(temp$dates)
  f <- which(x$dates %in% temp$dates)
  f_c <- which(names(x) %in% names(temp))
  
  x[f,f_c] <- temp
  
  M$google[[i]] <- x
}
names(M$google) <- stream_google

```



# some plots

```{r}

layout(matrix(1:6,3,2,byrow = FALSE))
col1 <- c('red3','blue3','chartreuse4','cyan4','darkgoldenrod3','darkmagenta')
d$dates <- as.Date(d$dates,format='%Y-%m-%d')
add_c <- c(FALSE,TRUE)

for(i in 2:ncol(d)){
  plot(d$dates,d[,i],main = names(d)[i])
  #apple
  c <- 1
  for(k in 1:length(stream_apple)){
    y <- M$apple[[k]][,i]
    
    if((sum(!is.na(y))>0) & (c==1)){
      plot(d$dates,y,pch=16,col=col1[k],main = 'apple',ylim = c(0,200))
      c <- 2
    }else if((sum(!is.na(y))>0) & (c==2)){
      lines(d$dates,y,type = 'p',pch=16,col=col1[k])
    }
  }
  if(c==1){
    plot(d$dates,rep(1e2,nrow(d)),ylim = c(0,1),main = 'apple')
  }
  
  #google
  c <- 1
  for(k in 1:length(stream_google)){
    y <- M$google[[k]][,i]
    if((sum(!is.na(y))>0) & (c==1)){
      plot(d$dates,y,pch=16,col=col1[k],main = 'google',ylim = c(0,200))
      c <- 2
    }else if((sum(!is.na(y))>0 ) & (c==2)){
      lines(d$dates,y,type = 'p',pch=16,col=col1[k])
    }
  }
  if(c==1){
    plot(d$dates,rep(1e2,nrow(d)),ylim = c(0,1),main = 'google')
  }
}


```









```{r}
#miss-spelled
country[-which(as.character(country) %in% as.character(d_m$region))]

# missing for apple: Algeria, China, Dominican_Republic, Ecuador, Iran, Peru, South_Korea, 
#corrections to match name in ECDC data
d_m$region[which(d_m$region %in% 'Czech Republic')] <- 'Czechia'
d_m$region[which(d_m$region %in% 'UK')] <- 'United_Kingdom'
d_m$region[which(d_m$region %in% 'United States')] <- 'United_States_of_America'
d_m$region[which(d_m$region %in% 'United Arab Emirates')] <- 'United_Arab_Emirates'
d_m$region[which(d_m$region %in% 'South Africa')] <- 'South_Africa'
d_m$region[which(d_m$region %in% 'Saudi Arabia')] <- 'Saudi_Arabia'
d_m$region[which(d_m$region %in% 'Republic of Korea')] <- 'South_Korea'
################

# same correction for google
d_m$region[which(d_m$region %in% 'United Kingdom')] <- 'United_Kingdom'
d_m$region[which(d_m$region %in% 'United States')] <- 'United_States_of_America'
d_m$region[which(d_m$region %in% 'South Korea')] <- 'South_Korea'
d_m$region[which(d_m$region %in% 'South Africa')] <- 'South_Africa'
d_m$region[which(d_m$region %in% 'Saudi Arabia')] <- 'Saudi_Arabia'
d_m$region[which(d_m$region %in% 'United Arab Emirates')] <- 'United_Arab_Emirates'
d_m$region[which(d_m$region %in% 'Dominican Republic')] <- 'Dominican_Republic'


#miss-spelled
country[-which(as.character(country) %in% as.character(d_m$region))]

# check
if (length(country[-which(country %in% d_m$region)])>0){
  print('warning')
  country[-which(country %in% d_m$region)]
}

# include country with data in both dataset
f1 <- which(country %in% d_m$region)
f2 <- which(d_m$region %in% country)

d_m <- d_m[f2,]

d$I_active_transmission <- d$I_active_transmission[,c(1,f1+1)]
d$D_active_transmission <- d$D_active_transmission[,c(1,f1+1)]
d$Country <- d$Country[f1]

# include dates with data in both
dates <- as.Date(substr(names(d_m[,-c(1:3)]),start = 2, stop = 11),format='%Y.%m.%d')
f1 <- which(d$I_active_transmission$dates %in% dates)
f2 <- which(dates %in% d$I_active_transmission$dates)

d_m <- d_m[,c(1,2,3,3+f2)]

d$I_active_transmission <- d$I_active_transmission[f1,]
d$D_active_transmission <- d$D_active_transmission[f1,]

```




# useful data

```{r}
# correct epi
D <- d$D_active_transmission
mD <- as.matrix(D[,-1])
I <- d$I_active_transmission

cumD <- data.frame(D$dates,apply(D[,-1],2,cumsum))
cumI <- data.frame(I$dates,apply(I[,-1],2,cumsum))

N_geo <- ncol(D)-1
N_days <- nrow(D)
country <- as.character(d$Country)

```


## process mobility data

load data (dots), get weekly average, interpolate between (blue line)

```{r, echo = FALSE}

mobility <- process_mobility(d_m = d_m)


layout(matrix(1:4,2,2))
for (i  in 1:N_geo){
  plot(mobility$mob_raw$dates,mobility$mob_raw[,i+1],main = country[i],ylim = c(0,1),
       xlab = '',ylab = 'prop. mobility')
  lines(mobility$mob_combined_smooth$dates,mobility$mob_combined_smooth[,i+1],col='blue')
  # lines(mob3[,i+1],col='red')
  abline(h=1,lwd=2,col='red',lty=2)
}
```



## process Epi data

load Epi data and serial interval
```{r, echo = FALSE}
## # serial interval estimate used: mean = 3.96, sd =  4.75
## # Teunis et al.

SI <- list()

for (si in c(1,2)){
  
  SI[[si]] <- list()
  SI[[si]]$si_mean <- d$si_mean[si]
  SI[[si]]$si_std <- d$si_std[si]
  
  SI[[si]]$SI <- gamma_dist_EpiEstim(si_mu =  d$si_mean[si], 
                            si_std  = d$si_std[si], 
                            SItrunc = 30)
  
  plot(seq(0, SI[[si]]$SI$SItrunc), SI[[si]]$SI$dist, type = "h",
       lwd = 10, lend = 1, xlab = "time (days)", ylab = "frequency")
  title(main = "Discrete distribution of the serial interval of COVID-19")
  
}

```

## delay infection to death

```{r, echo = FALSE}

delta_id <- gamma_dist_EpiEstim(si_mu = mu_id , 
                                si_std  = .45*mu_id, 
                                SItrunc = 50)

plot(seq(0, delta_id$SItrunc), delta_id$dist, type = "h",
     lwd = 10, lend = 1, xlab = "time (days)", ylab = "frequency")
title(main = "Discrete distribution of the delay infection to death")

sum(delta_id$dist[1:7])


```

# Useful formatting

see paper

```{r}

# delay death infection matrix
H <- matrix(0,nrow = N_days, ncol = N_days)
for (i in 1:N_days){
  f <- max(c(1,i-delta_id$SItrunc))
  H[i,f:i] <- rev(delta_id$dist)[((delta_id$SItrunc+1)-(i-f)):(delta_id$SItrunc+1)]
  if (i>1) H[i,f:i] <-  H[i,f:i]/sum( H[i,f:i])
}

# matrix of mobility and delayed version
M <- as.matrix(mobility$mob_combined_smooth[,-1])
# M_D <- H %*% M


# delay SI death to death matrix
W <- list()
Ot <- list()
for (si in c(1,2)){
  
  W[[si]] <- list()
  W[[si]]$W <- matrix(0,nrow = N_days, ncol = N_days)
  for (i in 1:N_days){
    f <- max(c(1,i-SI[[si]]$SI$SItrunc))
    W[[si]]$W[i,f:i] <- rev(SI[[si]]$SI$dist)[((SI[[si]]$SI$SItrunc+1)-(i-f)):(SI[[si]]$SI$SItrunc+1)]
    # if (i>1) H[i,f:i] <-  H[i,f:i]/sum( H[i,f:i])
  }
  
  # overall infectivity matrix
  Ot[[si]] <- list()
  Ot[[si]]$Ot <- W[[si]]$W %*% mD
  # correct infectivity when >0 case observed but no infectivity
  for(i in 1:N_geo){
     Ot[[si]]$Ot[ which( (D[,i+1]>0) & (Ot[[si]]$Ot[,i] ==0 ) ),i] <-NA
  }
}




```

# save output

```{r}

inputs <- list(D = D,
               M = M,
               Ot = Ot,
               W = W,
               H = H,
               SI = SI,
               delta_id = delta_id,
               mob_raw = mobility$mob_raw)

saveRDS(object = inputs, 
        file = paste0(path_s,'inputProcessing/Rdata/',
                      date_week_finishing,'_inputs_from_',D_source,'.rds'))

```